Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-20 reasoning.tex
edits

A central feature of a rely-guarantee judgment is a stability
condition that requires the validity of an assertion $\phi$ to be
unaffected by interference from other concurrently executing
transactions, i.e., the rely relation $R$. In conventional RG,
stability is defined as follows, where $\sigma$ and $\sigma'$ denote
states:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and the
availability of an isolation specification, we use multiple
definitions of stability in Fig.~\ref{fig:rg-rules}, but they all
convey the same intuition as above. In our setting, we only need to prove the stability of an assertion ($\phi$) against those environment steps which lead to a global database state on which the transaction itself can take its next step according to its isolation specification ($\I$).

\begin{smathpar}
\begin{array}{lcl}
\stable(R, \phi) & \Leftrightarrow & \forall \stl, \stg, \stg'. \phi(\stl, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow \phi(\stl, \stg')
\end{array}
\end{smathpar}

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if $\phi$
is preserved by every step of $R$, then $\phi$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of
a constrained relation. This means, unfortunately, that we cannot directly replace $R^{*}$
by $R$ in the above condition because of our need to consider the weak
isolation specification $\I$ that selectively exposes interferences.

To obtain an equivalent form in our setting, we require an additional
condition on the isolation specification, which we call the
\emph{stability condition on $\I$}.  The condition requires $\I$ to
allow interferences $R^{*}(\stg,\stg'')$, for two database states
$\stg$ and $\stg''$, only if it also allows interference for every
prefix of $R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows
interference from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to
$\stg''$ should not make the interference from $\stg$ to $\stg''$
valid.  This stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
modified a record, and no further interference will restore the
original record, because the original record bears the id of a
transaction that has already committed. We can now generalize the
stability condition for any arbitrary assertion $\phi$ by only
considering those environment steps which in addition to preserving
$\phi$, also preserve the condition that the transaction can continue
to take its next step, i.e. the condition $\I$ holds on both the
beginning and the end states.

 \begin{smathpar}
   \forall \stl, \stg, \stg', \stg_1. \phi(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg')
   \Rightarrow \phi(\stl, \stg')
\end{smathpar}

In the above condition, we use a separate notation ($\stg_1$) for the
global database state that the transaction witnessed during its last
step. $\phi$ does not constrain this global database state, but the
stability of $\I$ ensures that any database state constrained by $\phi$
must also obey $\I$. We can now replace the transitive closure in the
above condition with the single step relation $R$, backed by the
inductive argument for stability. Pushing the universal quantifier for
$\stg_1$ inside the antecedent and separating the predicates involving
$\stg_1$, we can rewrite the condition in the form below, by defining
a new ternary relation $R \backslash \I$, which can be thought of as
$R$ modulo $\I$:

\begin{smathpar}
\begin{array}{lcl}
\stable((R \backslash \I), \phi) & \Leftrightarrow & \forall \stl,\stg,\stg'. \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}

The above condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since if $\stg
\neq \stg'$, there does not exist any $\stg_1$ which satisfies the
constraint for $(R \backslash \I_{ss})(\stl,\stg,\stg')$.

@@

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It depends on a
transaction-local judgment to verify the body ($c$) of a transaction
with id $i$. The precondition $P$ of $c$ must follow from the fact
that the transaction-local database ($\stl$) is initially empty, and
the global database satisfies the high-level invariant $I$. The rely
relation ($\R_e$) is obtained from the global rely relation $R$ and the
isolation specification $\I_e$ as explained above. Recall that $\I_e$
constrains the global effects visible to the transaction while it is
executing but has not yet committed.

@@

Recall that $P$ and $Q$ of the transaction-local RG judgment are
binary assertions; they relate local and global database states. The
local judgment rules require one or both of them to be stable with
respect to the constrained rely relation $\R_e$.

@@

For the guarantee $G$ of a transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
$\R_e$, is computed as a rely relation modulo isolation, except that
commit-time isolation ($\I_c$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \rhd \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before a transaction
commits, then $G$ must relate the states of the global database before
and after the commit. The act of commit is captured by the flush
operation ($\stl\rhd\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

@@

\begin{proof}
Here, we provide only a brief proof sketch\footnote{The complete proof is provided in the supplementary material.}. The most important rule is the top-level rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which begins its execution in global database state satisfying $I$ and suffering interference $R$ while executing under isolation specification $\I$ finishes its execution in a database state also satisfying $I$, and also guarantees that its commit step satisfies $G$. The rule uses the transaction-local RG judgement $\R_e \vdash \rg{P}{c}{Q}$. By \rulelabel{E-Txn-Start}, the local and global database states at the start of a transaction satisfy $P$, and the only challenge is that environment steps in an execution covered by $\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level judgement requires environment steps in $R$. We show that it is enough to consider only those enviroment steps in $\R_e$. First, we use an inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to show that any execution in which the transaction completes all its steps must always preserve the isolation specification $\I_e$ after every environment step. Intuitively, this is because once $\I_e$ gets broken after some environment step, it will continue to remain broken and the transaction would not be able to proceed (according to \rulelabel{E-Txn}). Since $\R_e$ contains exactly those environment steps which preserve $\I_e$, the local-level RG judgement can be soundly used, which guarantees that after the transaction finishes its execution, its local state $\stl$ and global state $\stg$ will satisfy the assertion $Q$. Environment steps between the last step of the transaction and its commit step can modify the global state, and hence we also require $Q$ to be stable againt $R$. Again, we use an inductive argument, the stability of $\I_c$, and the fact that the transaction must execute its commit step to show that all environment steps must preserve $\I_c$, and hence it is enough to require $\stable(\R_c, Q)$. $Q$ guarantees that the commit step is in $G$, and $G$ in turn guarantees that after execution, the global database state will obey the invariant $I$.

The local RG-judgement rules are mostly straightforward and directly follow from the operational semantics. At this point, the transaction can only change its local database state by adding new records to it, and hence all the rules have the form : $P(\stl, \stg)$ and $\stl'$ is the new state should satisfy $Q(\stl', \stg)$. An important thing to note is that the rules for Update, Select, Delete and Insert only require the pre-condition $P$ to be stable against interference. This is because the local RG-judgement rules only make assertions about the database state just after the transaction takes its last step. For the aforementioned commands, the last step would be the only step taken by the transaction, and hence $Q$ would not be affected by any environment steps, but instead directly reflect the changes made by the command (the environment steps between the last execution step of the transaction and the commit step would be controlled by $\I_c$ and would be taken care of by $\stable(\R_c, Q)$ at the top level).

The rule $\rulelabel{RG-Foreach}$ makes use of an inductive loop invariant $\psi$ which assumes that the variable $y$ is bound to records which have been processed in previous iterations and $z$ is bound to the record processed in the current iteration. Since the operational semantics ensure that every iteration processes a unique record, the proof is straightforward.  
\end{proof}

>>>>

A central feature of a rely-guarantee judgment is a stability
condition that requires the validity of an assertion $\phi$ to be
unaffected by interference from other concurrently executing
transactions, i.e., the rely relation $R$. In conventional RG,
stability is defined as follows, where $\sigma$ and $\sigma'$ denote
states:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and the
availability of an isolation specification, we use multiple
definitions of stability in Fig.~\ref{fig:rg-rules}, but they all
convey the same intuition as above. In our setting, we only need to prove the stability of an assertion ($\phi$) against those environment steps which lead to a global database state on which the transaction itself can take its next step according to its isolation specification ($\I$). 
\begin{smathpar}
\begin{array}{lcl}
\stable(R, \phi) & \Leftrightarrow & \forall \stl, \stg, \stg'. \phi(\stl, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow \phi(\stl, \stg')
\end{array}
\end{smathpar}
\noindent A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if $\phi$
is preserved by every step of $R$, then $\phi$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of
a constrained relation. This means, unfortunately, that we cannot directly replace $R^{*}$
by $R$ in the above condition because of our need to consider the weak
isolation specification $\I$ that selectively exposes interferences.

To obtain an equivalent form in our setting, we require an additional
condition on the isolation specification, which we call the
\emph{stability condition on $\I$}.  The condition requires $\I$ to
allow interferences $R^{*}(\stg,\stg'')$, for two database states
$\stg$ and $\stg''$, only if it also allows interference for every
prefix of $R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows
interference from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to
$\stg''$ should not make the interference from $\stg$ to $\stg''$
valid.  This stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
modified a record, and no further interference will restore the
original record, because the original record bears the id of a
transaction that has already committed. We can now generalize the
stability condition for any arbitrary assertion $\phi$ by only
considering those environment steps which in addition to preserving
$\phi$, also preserve the condition that the transaction can continue
to take its next step, i.e. the condition $\I$ holds on both the
beginning and the end states.
 \begin{smathpar}
   \forall \stl, \stg, \stg', \stg_1. \phi(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg')
   \Rightarrow \phi(\stl, \stg')
\end{smathpar}
Here, we use a separate notation ($\stg_1$) for the global database
state that the transaction witnessed during its last step. $\phi$ does
not constrain this global database state, but the stability of $\I$
ensures that any database state constrained by $\phi$ must also obey
$\I$. We can now replace the transitive closure in the rule with the
single step relation $R$, backed by the inductive argument for
stability. Pushing the universal quantifier for $\stg_1$ inside the
antecedent and separating the predicates involving $\stg_1$, we can
rewrite the condition in the form below, by defining a new ternary
relation $R \backslash \I$, which can be thought of as $R$ modulo
$\I$:
\begin{smathpar}
\begin{array}{lcl}
\stable((R \backslash \I), \phi) & \Leftrightarrow & \forall \stl,\stg,\stg'. \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
This condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since if $\stg
\neq \stg'$, there does not exist any $\stg_1$ which satisfies the
constraint for $(R \backslash \I_{ss})(\stl,\stg,\stg')$.

@@

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It depends on a
transaction-local judgment to verify the body ($c$) of a transaction
with id $i$. The precondition $P$ of $c$ must follow from the fact
that the transaction-local database ($\stl$) is initially empty, and
the global database satisfies the high-level invariant $I$. The rely
relation ($\R_e$) is obtained from the global rely relation $R$ and
the isolation specification $\I_e$ as explained above. Recall that
$\I_e$ constrains the global effects visible to the transaction while
it is executing but has not yet committed, and $P$ and $Q$ of the
transaction-local RG judgment are binary assertions; they relate local
and global database states. The local judgment rules require one or
both of them to be stable with respect to the constrained rely
relation $\R_e$.

@@

@@

For the guarantee $G$ of a transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
$\R_e$, is computed as a rely relation modulo isolation, except that
commit-time isolation ($\I_c$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \rhd \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before a transaction
commits, then $G$ must relate the states of the global database before
and after the commit. The act of commit is captured by the flush
action ($\stl\rhd\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

@@

\paragraph{{\sc Proof Sketch.}}  The most important rule is the top-level
rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which
begins its execution in global database state satisfying $I$ and
encountering interference $R$ while executing under isolation
specification $\I$ finishes its execution in a database state also
satisfying $I$, and also guarantees that its commit step satisfies
$G$. The rule uses the transaction-local RG judgement $\R_e \vdash
\rg{P}{c}{Q}$. By \rulelabel{E-Txn-Start}, the local and global
database states at the start of a transaction satisfy $P$, and the
only challenge is that environment steps in an execution covered by
$\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level
judgement requires environment steps in $R$. We show that it is enough
to consider only those enviroment steps in $\R_e$. First, we use an
inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to
show that any execution in which the transaction completes all its
steps must always preserve the isolation specification $\I_e$ after
every environment step. Intuitively, this is because once $\I_e$ gets
broken after some environment step, it will continue to remain broken
and the transaction would not be able to proceed (according to
\rulelabel{E-Txn}). Since $\R_e$ contains exactly those environment
steps which preserve $\I_e$, the local-level RG judgement can be
soundly used, which guarantees that after the transaction finishes its
execution, its local state $\stl$ and global state $\stg$ will satisfy
the assertion $Q$. Environment steps between the last step of the
transaction and its commit step can modify the global state, and hence
we also require $Q$ to be stable againt $R$. Again, we use an
inductive argument, the stability of $\I_c$, and the fact that the
transaction must execute its commit step to show that all environment
steps must preserve $\I_c$, and hence it is enough to require
$\stable(\R_c, Q)$. $Q$ guarantees that the commit step is in $G$, and
$G$ in turn guarantees that after execution, the global database state
will obey the invariant $I$.

%% The local RG-judgement rules are mostly straightforward and directly
%% follow from the operational semantics. At this point, the transaction
%% can only change its local database state by adding new records to it,
%% and hence all the rules have the form : $P(\stl, \stg)$ and $\stl'$ is
%% the new state should satisfy $Q(\stl', \stg)$. Note that note is that
%% the {\sc RG-Update}, {\sf RG-Select}, {\sc RG-Delete} and {\sc
%%   RG-Insert} rules only require the pre-condition $P$ to be stable
%% against interference. This is because the these rules only make
%% assertions about the database state just after the transaction takes
%% its last step. For the aforementioned commands, the last step would be
%% the only step taken by the transaction, and hence $Q$ would not be
%% affected by any environment steps, but instead directly reflect the
%% changes made by the command (the environment steps between the last
%% execution step of the transaction and the commit step would be
%% controlled by $\I_c$ and would be taken care of by $\stable(\R_c, Q)$
%% at the top level).

%% The rule $\rulelabel{RG-Foreach}$ makes use of an inductive loop
%% invariant $\psi$ which assumes that the variable $y$ is bound to
%% records which have been processed in previous iterations and $z$ is
%% bound to the record processed in the current iteration. Since the
%% operational semantics ensure that every iteration processes a unique
%% record, the proof is straightforward.
